---
output: hugodown::hugo_document
slug: processx-3-9-0
title: processx 3.9.0
date: 2026-04-27
author: Gábor Csárdi
description: >
    processx 3.9.0 brings kernel-level process pipelines, pseudo-terminal
    support on Windows, Linux parent-death signals, binary I/O, and several
    other quality-of-life improvements.

photo:
  url: https://unsplash.com/photos/scUBcasSvbE
  author: Samuel Sianipar

# one of: "deep-dive", "learn", "package", "programming", "roundup", or "other"
categories: [package]
tags: [package, processx, processes, system]
---

<!--
TODO:
* [x] Look over / edit the post's title in the yaml
* [x] Edit (or delete) the description; note this appears in the Twitter card
* [x] Pick category and tags (see existing with `hugodown::tidy_show_meta()`)
* [x] Find photo & update yaml metadata
* [x] Create `thumbnail-sq.jpg`; height and width should be equal
* [x] Create `thumbnail-wd.jpg`; width should be >5x height
* [x] `hugodown::use_tidy_thumbnails()`
* [x] Add intro sentence, e.g. the standard tagline for the package
* [x] ~~`usethis::use_tidy_thanks()`~~
-->

We're happy to announce the release of
[processx](https://processx.r-lib.org/) 3.9.0. processx is an R package to
run and manage system processes.

You can install it from CRAN with:

```{r, eval = FALSE}
install.packages("processx")
```

This blog post discusses the major new features in processx 3.9.0.
You can see a full list of changes in the
[release notes](https://github.com/r-lib/processx/releases/tag/v3.9.0).

```{r setup, include = FALSE}
library(processx)
```

## Pipelines

New new `pipeline` class lets you connect two or more processes with
kernel-level pipes, exactly like a Unix shell pipeline
(`cmd1 | cmd2 | cmd3`): data flows directly between child processes without
passing through R.

```{r pipeline}
pl <- pipeline$new(
  list(c("sort"), c("uniq", "-c"), c("sort", "-rn")),
  stdin = "|",
  stdout = "|"
)
pl$write_input("banana\napple\nbanana\norange\napple\nbanana\n")
pl$close_input()
pl$read_all_output_lines()
pl$wait()
pl$get_exit_statuses()
```

The `pipeline$new()` constructor takes a list of character vectors — one
per command — along with the usual `stdin`, `stdout`, and `stderr`
arguments. These apply to the *ends* of the pipeline: `stdin` connects to
the first process, `stdout` reads from the last, and `stderr` controls all
processes.

The key benefit over calling `run()` in sequence is efficiency:
intermediate data never materialises in R. A pipeline processing gigabytes
of log lines uses the same small kernel buffers as a shell pipeline would.

Because each step in the pipeline is a regular `process` object under the
hood, you can access individual processes via `$get_processes()` — useful
for reading per-process stderr or checking exit codes when a stage fails.

`pipeline` works on Unix and Windows and is currently experimental: the
API may still change.

## Pseudo-terminal support

### `processx::run(pty = TRUE)`

Many command-line tools behave differently when their output is not
connected to a terminal: they disable colour, turn off progress bars,
or buffer output more aggressively.
The `pty = TRUE` option runs a process inside a pseudo-terminal so it
sees a real terminal — colour and interactive behaviour included.

`run()` now supports `pty = TRUE` directly:

```{r pty-run}
out <- run("ls", c("--color", path.expand("~/works/processx")), pty = TRUE)
cat(out$stdout)
```

When `pty = TRUE`, stderr is merged into stdout (the result's `$stderr` is
always `NULL`), because a PTY has a single stream. You can also supply a
file path as `stdin`; its contents are fed to the process via the PTY
master, followed by an EOF signal.

### Windows support

processx 3.9.0 adds support for pseudo-terminals (PTYs) on Windows,
starting from Windows 10 version 1809. The Windows implementation uses the
ConPTY API (`CreatePseudoConsole`), loaded dynamically so processx
continues to load on older Windows and emits a clear error if `pty = TRUE`
is requested on an unsupported version.

## Other improvements

### New process cleanup article

A new article,
[Process cleanup](https://processx.r-lib.org/articles/cleanup.html),
documents all five mechanisms processx provides for ensuring subprocesses
don't outlive their intended scope:

1. Explicit cleanup with `on.exit()` — always deterministic.
2. Automatic cleanup on garbage collection (`cleanup = TRUE`, the default).
3. Process-tree cleanup (`cleanup_tree = TRUE`).
4. Linux parent-death signal (`linux_pdeathsig`) — Linux only, handles R
   crashes.
5. Supervisor process (`supervise = TRUE`) — all platforms, handles R
   crashes.

### Death signal support on Linux

On Linux, you can now tell the kernel to deliver a signal to the child
process automatically if the parent R process exits — even if R crashes.
Set `linux_pdeathsig = TRUE` to send `SIGTERM`, or pass an integer signal
number directly:

```r
p <- process$new("sleep", "100", linux_pdeathsig = TRUE)
```

This is useful when you want child processes to clean up after an R crash,
without the overhead of running a supervisor. The argument is silently
ignored on macOS and Windows.

### Record the time when a process exits

`process$get_end_time()` returns the time when the process exited as a
`POSIXct`, or `NULL` if it is still running. This makes it straightforward
to measure wall-clock duration without having to record timestamps
yourself:

```{r end-time}
p <- process$new("sleep", "1")
p$wait()
p$get_end_time() - p$get_start_time()
```

### Append stdout/stderr to files

`process$new()` and `run()` now support `">>"` as a prefix for `stdout`
and `stderr` file paths to append output instead of truncating the file:

```{r append}
log <- tempfile()
run("echo", args = "first line", stdout = log)
run("echo", args = "second line", stdout = paste0(">>", log))
readLines(log)
```

This is handy when you run the same process repeatedly and want to
accumulate output in a single log file.

### Binary standard output and error

`run()` and `process$new()` now support `encoding = "binary"` to capture
raw bytes. In binary mode, `run()` returns `stdout` and `stderr` as raw
vectors, and `process$read_output()` / `process$read_error()` return raw
vectors rather than character strings. All bytes are preserved exactly,
including null bytes and non-UTF-8 sequences.

```{r binary}
result <- run("cat", args = "/bin/ls", encoding = "binary")
typeof(result$stdout)
length(result$stdout)
```

Two new methods, `process$read_output_bytes()` and
`process$read_error_bytes()`, and the `conn_read_bytes()` function,
provide direct access to raw bytes from processx connections.

## Acknowledgements

Thanks to everyone who contributed to processx 3.9.0 through code, issues,
testing, and feedback:

[@advieser](https://github.com/advieser),
[@cderv](https://github.com/cderv),
[@chwpearse](https://github.com/chwpearse),
[@HenrikBengtsson](https://github.com/HenrikBengtsson),
[@king-of-poppk](https://github.com/king-of-poppk),
[@r2evans](https://github.com/r2evans),
[@sckott](https://github.com/sckott),
[@sda030](https://github.com/sda030),
[@stupidpupil](https://github.com/stupidpupil), and
[@Yunuuuu](https://github.com/Yunuuuu).
